Accelerating Metagenomic Identification of DNA Sequences Using Artificial Neural Networks

### 2025-05-06 article/main.tex
article: description for results section

    \abstract{\todo{do przejrzenia, nie wszedzie poprawiony}The rapid increase in the availability of DNA sequence data necessitates efficient and scalable methods for metagenomic identification. Traditional tools like BLAST, although accurate, often require significant computational resources and time, limiting their applicability to large datasets. Recent advances in artificial neural networks (ANNs) demonstrate their potential to address this issue by offering faster and more accurate solutions.

@@

                \todo{wstep do badan, cel - zwiekszenie wydajnosci przy zachowaniu jakosci;
                  badania wymagaly 2 metod klasycznych; byly przeprowadzone na tych samych danych;}

@@

                \todo{tutaj metody klasyczne, porownawcze}

@@

            \subsubsection{Modified Needleman-Wunshc Algorithm}
                The first classical approach is the Needleman-Wunsch algorithm. This algorithm allows for global alignment of genetic sequences. It uses a similarity matrix between sequences, which is constructed according to the rules given in equation~\eqref{Equation:NeedlemanWunsch}. A modification of the approach presented in equation~\eqref{Equation:NeedlemanWunschModified} is applied in the work.

@@

        \subsection{Results and Analysis}
            \subsubsection{Experiment 1: Metagenomic identification execution time}
                \paragraph{Objective} 
                Measuring the execution time of metagenomic identification using different genetic sequence clustering methods.

@@

                \begin{table}\centering
                    \caption{Metagenomic identification execution time.}\label{Table:Experiment:Duration}
                    \begin{tabular}{|c||c|c|c|c|}
                        \hline
                        \multirow{2}{*}{\textbf{Number of sequences}} & \multicolumn{4}{|c|}{\textbf{Execution time [s]}} \\
                                        & \textbf{NP} & \textbf{NW} & \textbf{$k$-mer} & \textbf{ANN} \\ \hline \hline
                                        1 & 7107 & \textbf{6025} & 6087 & 6077\\ \hline
                                        2 & 6129 & 5994 & \textbf{5988} & 6035\\ \hline
                                        4 & 6108 & \textbf{5983} & 6024 & 6019\\ \hline
                                        8 & 6196 & 5988 & \textbf{5925} & 6014\\ \hline
                                        16 & 6290 & \textbf{5941} & 5946 & 6118\\ \hline
                                        32 & 6279 & 6035 & \textbf{5962} & 6092\\ \hline
                                        64 & 6316 & 6113 & \textbf{6107} & 6139\\ \hline
                                        128 & 6560 & 6238 & \textbf{6206} & 6233\\ \hline
                                        256 & 6753 & 6446 & \textbf{6295} & 6363\\ \hline
                                        512 & 7091 & 6972 & \textbf{6326} & 6347\\ \hline
                                        1024 & 8059 & 8743 & \textbf{6248} & 6269\\ \hline
                                        2048 & 8517 & 16461 & 6472 & \textbf{6332}\\ \hline
                                        4096 & 9433 & \textbf{-1} & 7003 & 6550\\ \hline

>>>>

    \abstract{The rapid increase in the availability of DNA sequence data necessitates efficient and scalable methods for metagenomic identification. Traditional tools like BLAST, although accurate, often require significant computational resources and time, limiting their applicability to large datasets. Recent advances in artificial neural networks (ANNs) demonstrate their potential to address this issue by offering faster and more accurate solutions.

@@

                This sections presents the results of the experiments conducted to evaluate the proposed solution, which aims to improve performance while maintaining the quality of metagenomic identifacation. The study involved comparing the new approach with two classical methods using the same dataset. The evaluation focused on execution time and identification quality to assess the effectiveness of the proposed approach.

@@

                \todo{tutaj metody klasyczne, porownawcze. PG: czy trzeba coś tutaj jeszcze dodawać?}

@@

            \subsubsection{Modified Needleman-Wunsch Algorithm}
                The first classical approach is the Needleman-Wunsch algorithm. This algorithm allows for global alignment of genetic sequences. It uses a similarity matrix between sequences, which is constructed according to the rules given in equation~\eqref{Equation:NeedlemanWunsch}. A modification of the approach presented in equation~\eqref{Equation:NeedlemanWunschModified} is applied in the work.

@@

        \subsection{Experiments}
            \subsubsection{Experiment 1: Metagenomic identification execution time}
                \paragraph{Objective} 
                Measuring the execution time of metagenomic identification using different genetic sequence clustering methods.

@@

                \begin{table}\centering
                    \caption{Metagenomic identification execution time.}\label{Table:Experiment:Duration}
                    \begin{tabular}{|c||c|c|c|c|}
                        \hline
                        \multirow{2}{*}{\textbf{Number of sequences}} & \multicolumn{4}{|c|}{\textbf{Execution time [s]}} \\
                                        & \textbf{NP} & \textbf{NW} & \textbf{$k$-mer} & \textbf{ANN} \\ \hline \hline
                                        1 & 7107 & \textbf{6025} & 6087 & 6077\\ \hline
                                        2 & 6129 & 5994 & \textbf{5988} & 6035\\ \hline
                                        4 & 6108 & \textbf{5983} & 6024 & 6019\\ \hline
                                        8 & 6196 & 5988 & \textbf{5925} & 6014\\ \hline
                                        16 & 6290 & \textbf{5941} & 5946 & 6118\\ \hline
                                        32 & 6279 & 6035 & \textbf{5962} & 6092\\ \hline
                                        64 & 6316 & 6113 & \textbf{6107} & 6139\\ \hline
                                        128 & 6560 & 6238 & \textbf{6206} & 6233\\ \hline
                                        256 & 6753 & 6446 & \textbf{6295} & 6363\\ \hline
                                        512 & 7091 & 6972 & \textbf{6326} & 6347\\ \hline
                                        1024 & 8059 & 8743 & \textbf{6248} & 6269\\ \hline
                                        2048 & 8517 & 16461 & 6472 & \textbf{6332}\\ \hline
                                        4096 & 9433 & \textbf{timeout} & 7003 & 6550\\ \hline
